Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer

### 2019-01-28 Paper/Paper.tex
replies to comments

%% The following should allow me to make subsubsubsections with the command \paragraph{}
\usepackage{titlesec}

@@

\section{Introduction}
Immunotherapy is slowly revolutionizing how we treat cancer.
Year-by-year, new breakthroughs in the field produce new, effective drugs and therapies that have a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
%A More examples? Reviews or non-reviews? Explain some of the examples?
There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area. Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
Understanding each step in this immune cascade in the context of the TME will be necessary in order to fully explore and exploit all options of immunotherapy.

@@

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
%A Should this be justified here or mentioned in the intro?
Upon successful clearance, they themselves are then deactivated and removed from the immune population.
%
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

@@

\subsubsection{Tissue cell fate}
During each cell cycle, every cell randomly is assigned a cell fate from the following options:
\begin{itemize}
\item proliferation
\item apoptosis
\item immune clearance (by NKs or CTLs)
\item rest/$G_0$ %A use both here?
\end{itemize}

@@

\paragraph{Rest/$G_0$} %A use both for this section heading?
The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
For Mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
Hence, the weight of quiescence is given by

$$ p_{G0} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$ 
%A style choice: p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???

The reason for adding that term is due to the understanding that mesenchymal cells are not proliferating as fast because they are instead remaining in the $G_0$ phase. %A is this the right way to word this?

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
%A Similar to previous comment: should I cite reference for this here or put this up in the intro?
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
%
Also, all proliferating cells have a probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
%A Updated this after Adam had question. Does this make sense as is?
If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. 
%
Then, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
Each cell receives an amount of TGF-$\beta$ given by

@@

%%%%%%%%%	I think this next section is too much detail that just confuses the reader. I don't think I need to explain my implementation, just the process. Thoughts?	%%%%%%%%%%%%%%%%%%%%
This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
If $\text{TGFB}_i>2-\text{EMT}_i$, the cell becomes more mesenchymal and will have a higher EMT score.
If $\text{TGFB}_i<2-\text{EMT}_i$, the cell becomes more epithelial and will have a lower EMT score.
The reason for the expression $2-\text{EMT}_i$ can be understood this way: 
the bounds on the EMT score guarantee $2-\text{EMT}_i\in[1,2]$.
Hence, there either needs to be a lot of TGF-$\beta$ in the system or a large amount of randomness for a cell to undergo EMT.
In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%% I like the first of these two options. You set aside the amount you know you will gain (sigma/d) and then let the rest decay exponentially
$$ \text{NK}_{k+1} = \left (\text{NK}_k-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$

@@

\subsection{Exploring the Model}\label{ExplModel}
In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
%A It seems the rest of this paragraph belongs in the Methods section, yeah?
Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
Any cell that has a pathway mutation is counted as a mutated cell and contributes towards the total number of cancer cells.
The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
In the model the EMT score is continuous but the labeling of epithelial or mesenchymal is a binary determined by being below or above a threshold.

%A This next commented out paragraph seems to belong somewhere, but where? I would think in the Intro.

@@

For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
The inflammation cycling scheme shown at the top of Figure \ref{fig:ModelIntro}B is repeated throughout.
We will return to this shortly.
It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, which translates to 405.75 days. %A translate to days like this?
At the same time as mutated cells are taking over, most cells are transitioning to a mesenchymal phenotype.
After the Time to Cancer, the mutated cells continue their rapid growth and soon make up 100\% of the cells.
Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.
% Once the Time to Cancer has been determined for a patient, most simulations shown later stop. [??? mean to say: we stop the simulation?]
%We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.

@@

\subsection{Sensitivity Analysis}\label{SensAnalysis}
To assess the sensitivity of the model to the 31 model parameters, a Morris OAT algorithm was implemented.
The algorithm works by running the model many successive times with exactly one parameter changing between each run\cite{morris1991factorial}\cite{sohier2014improvement}.
%A is the above description sufficient? The first source is Morris's original paper and the second is the source of the particular implementation I found
The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
Those that needed bounds were truncated.
The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}D.
Clearly, a certain subset of parameters have an outsized effect on the model while others do relatively little.
The two most influential according to this analysis are related to Treg cells: their death rate and recruitment.
This makes sense as Treg cells both suppress cytotoxic effects of other immune cells and also secrete significant quantities of TGF-$\beta$ which drives EMT.
Thus, Treg cells are intimately tied to all three aspects of the model and so their presence or lack thereof significantly affects the system.
However, we want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
In addition, inflammation parameters dictating the cycling scheme are worthy of study because of their prominence in Guo's work.
What we can see from this plot is that the model also has some high sensitivity to these parameters.
In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and so we will also look at that parameter.

@@

\subsection{Key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
Given the relevance of the inflammatory state of the tumor microenvironment (TME), we explore the effect of varying the inflammation state of the patient on survival.
In particular, some cohorts are assigned permanently low inflammation, others permanently high, and still others non-constant inflammatory schemes.
For those cohorts with a permanently high inflammatory state, the relationship between the mesenchymal parameters and the Time to Cancer is monotonic.
However, when there are periods of a low inflammatory state, then intriguingly the connection between MGA and Time to Cancer becomes concave down with a local maximum around 0.3.
This can be seen in the top two groups of Figure \ref{fig:VaryINFL_and_MesPars}D in which the MGA=0.3 cohorts both took longer to cancer than their neighbors.
What this indicates is that if the growth arrest of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer.
Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
%A Is this next sentence necessary?
See Figure \ref{fig:VaryINFL_and_MesPars}C and Figure \ref{fig:VaryINFL_and_MesPars}D for a summary.

@@

Contrast this with what happens when MIE is varied for different inflammation cycling schemes.
Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is always -monotonic.
In the case that the inflammatory state is permanently high, MIE has no significant effect on Time to Cancer.
On the other hand, when the patient experiences at least some time in a low inflammatory state, an increase in MIE results in a decrease in Time to Cancer.
This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
%A Is this next sentence necessary?
See Figure \ref{fig:VaryINFL_and_MesPars}A and Figure \ref{fig:VaryINFL_and_MesPars}B for a summary.

@@

The goal of improving cancer treatment has always proven attainable even despite great difficulty.
The processes involved are so complex, yet it is those complexities that we can exploit to better control and eradicate cancer.
Thus, continuing research into every aspect of cancer is the clearest, best path forward.

>>>>

%% The following should allow me to make subsubsubsections with the command \paragraph{}
%% AM: I don't think this is necessary. please use \paragraph (original style) instead
\usepackage{titlesec}

@@

\section{Introduction}
Immunotherapy is slowly revolutionizing how we treat cancer.
Year-by-year, new breakthroughs in the field produce new, effective drugs and therapies that have a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
% A More examples? Reviews or non-reviews? Explain some of the examples?
%AM I will add a few 
There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area. Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
Understanding each step in this immune cascade in the context of the TME will be necessary in order to fully explore and exploit all options of immunotherapy.

@@

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
% A Should this be justified here or mentioned in the intro?
% AM: justifed no; Intro yes
Upon successful clearance, they themselves are then deactivated and removed from the immune population.
%
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

@@

\subsubsection{Tissue cell fate}
During each cell cycle, every cell randomly is assigned a cell fate from the following options:
\begin{itemize}
\item proliferation
\item apoptosis
\item immune clearance (by NKs or CTLs)
\item rest in $G_0$ %A use both here? - % AM: change to 'rest in' everywhere
\end{itemize}

@@

\paragraph{Rest in $G_0$} %A use both for this section heading?
The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
For Mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
Hence, the weight of quiescence is given by

$$ p_{rest} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$ 
%YA style choice: p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???

The reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells remain longer in the $G_0$ phase. %YA is this the right way to word this?

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
% A Similar to previous comment: should I cite reference for this here or put this up in the intro?
% AM no need for extra here: just in intro
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
%
All proliferating cells have a nonzero probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
% A Updated this after Adam had question. Does this make sense as is?
% AM I don't understnad 'remains mutated'
If the cell does not undergo a mutation, then its probability of mutation during subsequent cell increases. 
%
Then, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
Each cell receives an amount of TGF-$\beta$ given by

@@

%%%%%%%%%	I think this next section is too much detail that just confuses the reader. I don't think I need to explain my implementation, just the process. Thoughts?	%%%%%%%%%%%%%%%%%%%%
% AM: yes, remove or shorten to one sentence at end of previous para.
This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
If $\text{TGFB}_i>2-\text{EMT}_i$, the cell becomes more mesenchymal and will have a higher EMT score.
If $\text{TGFB}_i<2-\text{EMT}_i$, the cell becomes more epithelial and will have a lower EMT score.
The reason for the expression $2-\text{EMT}_i$ can be understood this way: 
the bounds on the EMT score guarantee $2-\text{EMT}_i\in[1,2]$.
Hence, there either needs to be a lot of TGF-$\beta$ in the system or a large amount of randomness for a cell to undergo EMT.
In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%% I like the first of these two options. You set aside the amount you know you will gain (sigma/d) and then let the rest decay exponentially
% AM yes first one 
$$ \text{NK}_{k+1} = \left (\text{NK}_k-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$

@@

\subsection{Exploring the Model}\label{ExplModel}
In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
% A It seems the rest of this paragraph belongs in the Methods section, yeah?
% AM : yes
Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
Any cell that has a pathway mutation is counted as a mutated cell and contributes towards the total number of cancer cells.
The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
In the model the EMT score is continuous but the labeling of epithelial or mesenchymal is a binary determined by being below or above a threshold.

% A This next commented out paragraph seems to belong somewhere, but where? I would think in the Intro.
% AM: yes in intro

@@

For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
The inflammation cycling scheme shown at the top of Figure \ref{fig:ModelIntro}B is repeated throughout.
We will return to this shortly.
It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, or 405.75 days. 
At the same time as mutated cells are taking over, most cells are transitioning to a mesenchymal phenotype.
After the Time to Cancer, the mutated cells continue their rapid growth and soon make up 100\% of the cells.
Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.
% Once the Time to Cancer has been determined for a patient, most simulations shown later stop. [??? mean to say: we stop the simulation?]
%We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.

@@

\subsection{Sensitivity Analysis}\label{SensAnalysis}
To assess the sensitivity of the model to the 31 model parameters, a Morris OAT algorithm was implemented.
The algorithm works by running the model many successive times with exactly one parameter changing between each run\cite{morris1991factorial}\cite{sohier2014improvement}.
% A is the above description sufficient? The first source is Morris's original paper and the second is the source of the particular implementation I found
% AM: yes fine 
The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
Those that needed bounds were truncated.
The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}D.
Clearly, a certain subset of parameters have an outsized effect on the model while others do relatively little.
The two most influential according to this analysis are related to Treg cells: their death rate and recruitment.
This makes sense as Treg cells both suppress cytotoxic effects of other immune cells and also secrete significant quantities of TGF-$\beta$ which drives EMT.
Thus, Treg cells are intimately tied to all three aspects of the model and so their presence or lack thereof significantly affects the system.
However, we want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
In addition, inflammation parameters dictating the cycling scheme are worthy of study because of their prominence in Guo's work.
What we can see from this plot is that the model also has some high sensitivity to these parameters.
In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and so we will also look at that parameter.

@@

\subsection{Key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
Given the relevance of the inflammatory state of the tumor microenvironment (TME), we explore the effect of varying the inflammation state of the patient on survival.
In particular, some cohorts are assigned permanently low inflammation, others permanently high, and still others non-constant inflammatory schemes.
For those cohorts with a permanently high inflammatory state, the relationship between the mesenchymal parameters and the Time to Cancer is monotonic.
However, when there are periods of a low inflammatory state, then intriguingly the connection between MGA and Time to Cancer becomes concave down with a local maximum around 0.3.
This can be seen in the top two groups of Figure \ref{fig:VaryINFL_and_MesPars}D in which the MGA=0.3 cohorts both took longer to cancer than their neighbors.
What this indicates is that if the growth arrest of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer.
Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
% A Is this next sentence necessary?
% AM: can remove it 
See Figure \ref{fig:VaryINFL_and_MesPars}C and Figure \ref{fig:VaryINFL_and_MesPars}D for a summary.

@@

Contrast this with what happens when MIE is varied for different inflammation cycling schemes.
Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is always -monotonic.
In the case that the inflammatory state is permanently high, MIE has no significant effect on Time to Cancer.
On the other hand, when the patient experiences at least some time in a low inflammatory state, an increase in MIE results in a decrease in Time to Cancer.
This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
% A Is this next sentence necessary?
% AM: remove
See Figure \ref{fig:VaryINFL_and_MesPars}A and Figure \ref{fig:VaryINFL_and_MesPars}B for a summary.

@@

Progress in cancer treatment has always proven attainable even despite great challenges, for which the complexity of the disease is often responsible. As we move forwards, it is these very complexities that we will be able to better exploit to eradicate or control the disease.
